Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression

### 2018-12-02 paper/paper_no_event.tex
Paper: 2nd paragraph of intro + small edits.

The reliability of these measurements, however, depend on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data. Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. To avoid large systematic errors, highly accurate waveform models are needed that include all relevant physical effects such as point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Most time-domain waveform implementations involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain. Depending on the model, this can take seconds to hours for long BNS waveforms, so analytic frequency-domain approximations are often used instead.

For the analysis of GW170817 in Ref.~\cite{BNSPE}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models, a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

In this paper, we will use an implementation of the EOB formalism that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a different point-particle and spin prescription. For matter effects, this model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.

@@

\red{[check notations have been introduced]} For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins. It includes consistently spin-orbit interactions up to 3.5PN order and spin-spin interactions up to 2PN order. In addition, the \texttt{SEOBNRv4T} model accounts for the spin-induced quadrupole moment of neutron stars~\cite{Poisson1998}, an extended-body effect distinguishing neutron stars from black holes. This effect is quadratic in the spins, appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin~\cite{HarryHinderer2018}. Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only, see~\cite{}.}, at the moment the effect is only consistently included at the leading 2PN order. We leave for future work an extension to higher orders\footnote{\red{complete this} We note that TaylorF2 models do include all spin-square terms at 3PN, see App.~\ref{sec:taylorf2}. They are also present in the IMRPhenomPv2+NRTides model, but not in the SEOBNRv4ROM+NRTides model. TEOBResumS is also limited to the leading 2PN order.}.

@@

Given the GPR fits for the amplitude and phase residuals, we can now reconstruct the frequency-domain waveform. We label the set of GPR fits at the frequency nodes $MF_j$ by $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ for the amplitude residuals and by $\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}$ for the phase residuals. The surrogates for the residuals $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$ are constructed by interpolating between the nodes $MF_j$ with cubic splines:
\begin{align}
\Delta\ln A_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}](Mf),\\
\Delta\Phi_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}](Mf).
\end{align}
We set these functions to 0 below the first frequency node $MF_0 =  Mf_{{\rm trunc},i}$ so that the waveform transitions to TaylorF2 at lower frequencies. With the analytic expressions for $A_{\rm F2}(Mf;\bx)$ and $\Phi_{\rm F2}(Mf;\bx)$ and the interpolated expressions $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$, the final surrogates for the amplitude and phase are
\begin{align}
\label{eq:amp_sur}
A_S(Mf;\bx) &= A_{\rm F2}(Mf;\bx) \exp\left[ \Delta\ln A_S(Mf;\bx) \right], \\
\label{eq:phase_sur}
\Phi_S(Mf;\bx) &= \Phi_{\rm F2}(Mf;\bx) + \Delta\Phi_S(Mf;\bx).
\end{align}
In physical units, for an inclination angle $\iota$, the $+$ and $\times$ polarizations of the waveform are 
\begin{align}
\label{eq:hplus_sur}
\tilde h_+(f; \bx) &= \frac{1}{2}(1+\cos^2\iota) \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right)\right], \\
\label{eq:hcross_sur}
\tilde h_\times(f; \bx) &= \cos\iota \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right) + i \frac{\pi}{2}\right].
\end{align}

@@

We show marginalized 1-dimensional posteriors for four key parameters in Fig.~\ref{fig:1dpe}. These are the chirp mass $\mathcal{M} = (m_1 m_2)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$. As the SNR increases, the peaks of the marginalized 1-d posteriors become more closely aligned with the injected value. However, the distributions are noticeably asymmetric and are not centered on the injected values. 
This results mainly because prior boundaries on some parameters, such as the mass ratio $q$, shift the marginalized PDFs for correlated parameters.
This effect is strongest for the interaction between $\chi_{\rm eff}$ and $q$ (Fig.~\ref{fig:2dpe}), where the injected value $q=1$ is at the boundary of the prior. Although the injected parameter $(\chi_{\rm eff}, q) = (0.1, 1)$ lies on the ridge of maximum density in the marginalized 2-d PDF, the marginalized 1-d PDF for $\chi_{\rm eff}$ is noticeably offset. To counteract this effect in Fig.~\ref{fig:1dpe}, we also take a thin cross section containing the injected value of $q=1$ by cutting samples outside the interval $q \in [0.9, 1]$. This causes the peaks of the other 1-d distributions to be much closer to the injected values. We consider the difference between these peaks and the injected values to be a conservative bound on the size of the statistical errors due to errors in the surrogate model.

>>>>

The reliability of these measurements, however, depend on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data. Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. To avoid large systematic errors, highly accurate waveform models are needed that include all relevant physical effects such as point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Most time-domain waveform implementations involve solving a set of ordinary differential equations, which depending on the model can take seconds to hours for long BNS waveforms. The results then need to be transformed into the frequency domain where the analysis takes place. Because of this, analytic approximations providing waveforms directly in the frequency-domain  are often used instead.

For the analysis of GW170817 in Ref.~\cite{BNSPE}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either an aligned-spin post-Newtonian (PN) approximant in the Fourier domain (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models, a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

In this paper, we will use an implementation of the EOB formalism that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018, TEOBResumSupdate}, that uses a different point-particle and spin prescription. For matter effects, this model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.

@@

For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are based on the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins. It includes consistently spin-orbit interactions up to 3.5PN order and spin-spin interactions up to 2PN order. In addition, the \texttt{SEOBNRv4T} model accounts for the spin-induced quadrupole moment of neutron stars~\cite{Poisson1998}, an extended-body effect distinguishing neutron stars from black holes. This effect is quadratic in the spins, appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin~\cite{HarryHinderer2018}. Although complete PN expressions for this contribution in the dynamics and waveform are known at 3PN~\cite{Bohe:2015ana}\footnote{And at 4PN for the dynamics only, see~\cite{Levi:2015ixa}.}, at the moment the effect is only consistently included at the leading 2PN order. An extension to higher orders is left for future work\footnote{We note that TaylorF2 models do include all known spin-square terms at 3PN, see App.~\ref{sec:taylorf2}. They are also present in the IMRPhenomPv2+NRTides and SEOBNRv4ROM+NRTides models. TEOBResumS was recently updated~\cite{TEOBResumSupdate} to include these next-to-leading terms.}.

@@

Given the GPR fits for the amplitude and phase residuals, we can now reconstruct the frequency-domain waveform. We label the set of GPR fits at the frequency nodes $MF_j$ by $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ for the amplitude residuals and by $\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}$ for the phase residuals. The surrogates for the residuals $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$ are constructed by interpolating between the nodes $MF_j$ with cubic splines:
\begin{align}
\Delta\ln A_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}](Mf),\\
\Delta\Phi_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}](Mf).
\end{align}
We set these functions to 0 below the first frequency node $MF_0 =  Mf_{{\rm trunc},i}$ so that the waveform transitions to TaylorF2 at lower frequencies. With the analytic expressions for $A_{\rm F2}(Mf;\bx)$ and $\Phi_{\rm F2}(Mf;\bx)$ and the interpolated expressions $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$, the final surrogates for the amplitude and phase are
\begin{align}
\label{eq:amp_sur}
A_S(Mf;\bx) &= A_{\rm F2}(Mf;\bx) \exp\left[ \Delta\ln A_S(Mf;\bx) \right], \\
\label{eq:phase_sur}
\Phi_S(Mf;\bx) &= \Phi_{\rm F2}(Mf;\bx) + \Delta\Phi_S(Mf;\bx).
\end{align}
In physical units, for an inclination angle $\iota$, the $+$ and $\times$ polarizations of the waveform are (for $f>0$)
\begin{align}
\label{eq:hplus_sur}
\tilde h_+(f; \bx) &= \frac{1}{2}(1+\cos^2\iota) \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right)\right], \\
\label{eq:hcross_sur}
\tilde h_\times(f; \bx) &= \cos\iota \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right) + i \frac{\pi}{2}\right].
\end{align}

@@

We show marginalized 1-dimensional posteriors for four key parameters in Fig.~\ref{fig:1dpe}. These are the chirp mass $\mathcal{M} = (m_1 m_2)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$. As the SNR increases, the peaks of the marginalized 1-d posteriors become more closely aligned with the injected value. However, the distributions are noticeably asymmetric and are not centered on the injected values. 
This results mainly because prior boundaries on some parameters, such as the mass ratio $q$, shift the marginalized PDFs for correlated parameters.
This effect is strongest for the interaction between $\chi_{\rm eff}$ and $q$ (Fig.~\ref{fig:2dpe}), where the injected value $q=1$ is at the boundary of the prior. Although the injected parameter $(\chi_{\rm eff}, q) = (0.1, 1)$ lies on the ridge of maximum density in the marginalized 2-d PDF, the marginalized 1-d PDF for $\chi_{\rm eff}$ is noticeably offset. To counteract this effect in Fig.~\ref{fig:1dpe}, we also take a thin cross section containing the injected value of $q=1$ by cutting samples outside the interval $q \in [0.9, 1]$. This causes the peaks of the other 1-d distributions to be much closer to the injected values. We consider the difference between these peaks and the injected values to be a conservative bound on the size of the systematic errors due to errors in the surrogate model.
